---
title: "P8105 Homework 2"
author: "xj2357"
output: github_document
---

```{r setup, include = FALSE}
library(tidyverse)
```

# Problem 1

## Import the subway data

```{r import_subway}
subway_df =
  read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
  show_col_types = FALSE) |>
  janitor::clean_names() |>
  select(line, station_name, station_latitude, station_longitude,
  starts_with("route"), entry, vending, entrance_type, ada) |>
  mutate(entry = entry == "YES")

glimpse(subway_df)
```
The cleaned dataset contains `r nrow(subway_df)` entrance/exit records and `r ncol(subway_df)` variables. The variables describe the subway line, station name and coordinates, routes served, entry permission, vending availability, entrance type, and ADA compliance. 
I standardized the column names, retained the requested variables, and converted entry from YES/NO to a logical variable. The route information still needs tidying because it is spread across route1 through route11 rather than stored in a single route-name column.

## Station counts and entry access

```{r subway_questions}
n_stations =
  subway_df |>
  distinct(line, station_name) |>
  nrow()

n_ada_stations =
  subway_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()

entry_prop_no_vending =
  subway_df |>
  filter(vending == "NO") |>
  pull(entry) |>
  mean()
```

Stations are identified by the combination of line and station name.
There are `r n_stations` distinct stations, of which
`r n_ada_stations` are ADA compliant.

Among entrances/exits without vending, approximately
`r round(entry_prop_no_vending * 100, 2)`% allow entry.

## Tidy the route information

```{r tidy_subway_routes}
subway_long_df =
  subway_df |>
  mutate(
    across(starts_with("route"), as.character)
  ) |>
  pivot_longer(
    cols = starts_with("route"),
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name",
    values_drop_na = TRUE
  ) |>
  mutate(
    route_number = as.integer(route_number)
  )

n_a_stations =
  subway_long_df |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()

n_a_ada_stations =
  subway_long_df |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```

I converted the route columns to character variables and reshaped them
into route_number and route_name. Missing route entries were omitted.
Each row now represents one route served by an entrance/exit record.
The resulting dataset contains `r nrow(subway_long_df)` rows and
`r ncol(subway_long_df)` columns.

There are `r n_a_stations` distinct stations serving the A train,
of which `r n_a_ada_stations` are ADA compliant.

# Problem 2
## Import and clean Mr. Trash Wheel data

```{r import_mr_trash}
mr_trash_df =
  readxl::read_excel(
    "data/trash_wheel_202610.xlsx",
    sheet = "Mr. Trash Wheel",
    range = readxl::cell_limits(
      c(2, 1),
      c(NA, 14)
    )
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    year = as.integer(year),
    date = as.Date(date),
    sports_balls = as.integer(round(sports_balls)),
    wheel = "Mr. Trash Wheel"
  )

glimpse(mr_trash_df)
```
## Import Professor Trash Wheel and Gwynnda

```{r import_other_wheels}
professor_df =
  readxl::read_excel(
    "data/trash_wheel_202610.xlsx",
    sheet = "Professor Trash Wheel",
    range = readxl::cell_limits(c(2, 1), c(NA, 13)),
    na = c("", "dive")
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    year = as.integer(year),
    date = as.Date(date),
    wheel = "Professor Trash Wheel"
  )

gwynnda_df =
  readxl::read_excel(
    "data/trash_wheel_202610.xlsx",
    sheet = "Gwynnda the Good Wheel of the W",
    range = readxl::cell_limits(c(2, 1), c(NA, 12))
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    year = as.integer(year),
    date = as.Date(date),
    wheel = "Gwynnda"
  )
```

The Professor Trash Wheel worksheet contains the nonnumeric text "dive"
in one plastic-bottle count cell. I treated this value as missing while
retaining the record and its other available measurements.

## Combine the three datasets

```{r combine_trash_wheels}
trash_df =
  bind_rows(mr_trash_df, professor_df, gwynnda_df)

count(trash_df, wheel)

dim(trash_df)
```
## Summary of the combined dataset

```{r trash_summary}
professor_total_weight =
  trash_df |>
  filter(wheel == "Professor Trash Wheel") |>
  pull(weight_tons) |>
  sum(na.rm = TRUE)

gwynnda_june_2022_butts =
  trash_df |>
  filter(
    wheel == "Gwynnda",
    year == 2022,
    month == "June"
  ) |>
  pull(cigarette_butts) |>
  sum(na.rm = TRUE)
```

The combined dataset contains `r nrow(trash_df)` dumpster collection
records and `r ncol(trash_df)` variables from three trash wheels.
Key variables include collection date, trash weight in tons, volume
in cubic yards, and counts of items such as plastic bottles and
cigarette butts. The wheel variable identifies the source of each record.
I selected the relevant Excel ranges, standardized column names and
data types, removed rows without dumpster numbers, and combined the
three datasets. Sports-ball counts were rounded and stored as integers.
Variables absent from a source worksheet are recorded as missing
after combining.

Across the available records, Professor Trash Wheel collected
`r round(professor_total_weight, 2)` tons of trash.
Gwynnda collected `r gwynnda_june_2022_butts` cigarette butts
in June 2022.

# Problem 3