---
title: "P8105 Homework 2"
author: "xj2357"
output: github_document
---

```{r setup, include = FALSE}
library(tidyverse)
```

# Problem 1

## Import the subway data

```{r import_subway}
subway_df =
  read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
  show_col_types = FALSE) |>
  janitor::clean_names() |>
  select(line, station_name, station_latitude, station_longitude,
  starts_with("route"), entry, vending, entrance_type, ada) |>
  mutate(entry = entry == "YES")

glimpse(subway_df)
```
The cleaned dataset contains `r nrow(subway_df)` entrance/exit records and `r ncol(subway_df)` variables. The variables describe the subway line, station name and coordinates, routes served, entry permission, vending availability, entrance type, and ADA compliance. 
I standardized the column names, retained the requested variables, and converted entry from YES/NO to a logical variable. The route information still needs tidying because it is spread across route1 through route11 rather than stored in a single route-name column.

# Problem 2

# Problem 3