---
title: "P8105 Homework 2"
author: "xj2357"
output: github_document
---

```{r setup, include = FALSE}
library(tidyverse)
```

# Problem 1

## Import the subway data

```{r import_subway}
subway_df =
  read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
  show_col_types = FALSE) |>
  janitor::clean_names() |>
  select(line, station_name, station_latitude, station_longitude,
  starts_with("route"), entry, vending, entrance_type, ada) |>
  mutate(entry = entry == "YES")

glimpse(subway_df)
```
The cleaned dataset contains `r nrow(subway_df)` entrance/exit records and `r ncol(subway_df)` variables. The variables describe the subway line, station name and coordinates, routes served, entry permission, vending availability, entrance type, and ADA compliance. 
I standardized the column names, retained the requested variables, and converted entry from YES/NO to a logical variable. The route information still needs tidying because it is spread across route1 through route11 rather than stored in a single route-name column.

## Station counts and entry access

```{r subway_questions}
n_stations =
  subway_df |>
  distinct(line, station_name) |>
  nrow()

n_ada_stations =
  subway_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()

entry_prop_no_vending =
  subway_df |>
  filter(vending == "NO") |>
  pull(entry) |>
  mean()
```

Stations are identified by the combination of line and station name.
There are `r n_stations` distinct stations, of which
`r n_ada_stations` are ADA compliant.

Among entrances/exits without vending, approximately
`r round(entry_prop_no_vending * 100, 2)`% allow entry.

## Tidy the route information

```{r tidy_subway_routes}
subway_long_df =
  subway_df |>
  mutate(
    across(starts_with("route"), as.character)
  ) |>
  pivot_longer(
    cols = starts_with("route"),
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name",
    values_drop_na = TRUE
  ) |>
  mutate(
    route_number = as.integer(route_number)
  )

n_a_stations =
  subway_long_df |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()

n_a_ada_stations =
  subway_long_df |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```

I converted the route columns to character variables and reshaped them
into route_number and route_name. Missing route entries were omitted.
Each row now represents one route served by an entrance/exit record.
The resulting dataset contains `r nrow(subway_long_df)` rows and
`r ncol(subway_long_df)` columns.

There are `r n_a_stations` distinct stations serving the A train,
of which `r n_a_ada_stations` are ADA compliant.

# Problem 2
## Import and clean Mr. Trash Wheel data

```{r import_mr_trash}
mr_trash_df =
  readxl::read_excel(
    "data/trash_wheel_202610.xlsx",
    sheet = "Mr. Trash Wheel",
    range = readxl::cell_limits(
      c(2, 1),
      c(NA, 14)
    )
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    year = as.integer(year),
    date = as.Date(date),
    sports_balls = as.integer(round(sports_balls)),
    wheel = "Mr. Trash Wheel"
  )

glimpse(mr_trash_df)
```
## Import Professor Trash Wheel and Gwynnda

```{r import_other_wheels}
professor_df =
  readxl::read_excel(
    "data/trash_wheel_202610.xlsx",
    sheet = "Professor Trash Wheel",
    range = readxl::cell_limits(c(2, 1), c(NA, 13)),
    na = c("", "dive")
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    year = as.integer(year),
    date = as.Date(date),
    wheel = "Professor Trash Wheel"
  )

gwynnda_df =
  readxl::read_excel(
    "data/trash_wheel_202610.xlsx",
    sheet = "Gwynnda the Good Wheel of the W",
    range = readxl::cell_limits(c(2, 1), c(NA, 12))
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    year = as.integer(year),
    date = as.Date(date),
    wheel = "Gwynnda"
  )
```

The Professor Trash Wheel worksheet contains the nonnumeric text "dive"
in one plastic-bottle count cell. I treated this value as missing while
retaining the record and its other available measurements.

## Combine the three datasets

```{r combine_trash_wheels}
trash_df =
  bind_rows(mr_trash_df, professor_df, gwynnda_df)

count(trash_df, wheel)

dim(trash_df)
```
## Summary of the combined dataset

```{r trash_summary}
professor_total_weight =
  trash_df |>
  filter(wheel == "Professor Trash Wheel") |>
  pull(weight_tons) |>
  sum(na.rm = TRUE)

gwynnda_june_2022_butts =
  trash_df |>
  filter(
    wheel == "Gwynnda",
    year == 2022,
    month == "June"
  ) |>
  pull(cigarette_butts) |>
  sum(na.rm = TRUE)
```

The combined dataset contains `r nrow(trash_df)` dumpster collection
records and `r ncol(trash_df)` variables from three trash wheels.
Key variables include collection date, trash weight in tons, volume
in cubic yards, and counts of items such as plastic bottles and
cigarette butts. The wheel variable identifies the source of each record.
I selected the relevant Excel ranges, standardized column names and
data types, removed rows without dumpster numbers, and combined the
three datasets. Sports-ball counts were rounded and stored as integers.
Variables absent from a source worksheet are recorded as missing
after combining.

Across the available records, Professor Trash Wheel collected
`r round(professor_total_weight, 2)` tons of trash.
Gwynnda collected `r format(gwynnda_june_2022_butts, scientific = FALSE, big.mark = ",")` 
cigarette butts in June 2022.

# Problem 3

## Import and clean the baseline data

```{r import_baseline}
baseline_raw =
  read_csv(
    "data/data_mci/MCI_baseline.csv",
    skip = 1,
    na = c("", "NA", "."),
    show_col_types = FALSE
  ) |>
  janitor::clean_names()

baseline_df =
  baseline_raw |>
  rename(baseline_age = current_age) |>
  mutate(
    sex = factor(
      sex,
      levels = c(0, 1),
      labels = c("Female", "Male")
    ),
    apoe4 = factor(
      apoe4,
      levels = c(0, 1),
      labels = c("Non-carrier", "Carrier")
    )
  ) |>
  filter(
    is.na(age_at_onset) | age_at_onset > baseline_age
  )

nrow(baseline_raw)
nrow(baseline_df)
glimpse(baseline_df)
```

## Summarize the baseline data

```{r summarize_baseline}
n_mci =
  baseline_df |>
  filter(!is.na(age_at_onset)) |>
  nrow()

mean_baseline_age =
  baseline_df |>
  pull(baseline_age) |>
  mean()

women_apoe4_prop =
  baseline_df |>
  filter(sex == "Female") |>
  mutate(is_carrier = apoe4 == "Carrier") |>
  pull(is_carrier) |>
  mean()
```

I skipped the first row containing variable descriptions and treated "." as
missing. I standardized column names, renamed current_age to baseline_age,
and converted sex and APOE4 carrier status to factors. Participants whose
MCI onset age was at or before their baseline age were excluded. Missing
onset ages were retained because they indicate that participants remained
MCI-free during follow-up.

The original baseline dataset contains `r nrow(baseline_raw)` participants.
After excluding `r nrow(baseline_raw) - nrow(baseline_df)` participants who
already had MCI at baseline, `r nrow(baseline_df)` eligible participants
remain. The cleaned dataset has one row per participant and
`r ncol(baseline_df)` variables: participant ID, baseline age, sex, years of
education, APOE4 carrier status, and age at MCI onset.

Among the eligible participants, `r n_mci` developed MCI during follow-up.
Their average baseline age was `r round(mean_baseline_age, 2)` years.
Among women in this eligible cohort,
`r round(women_apoe4_prop * 100, 1)`% were APOE4 carriers.

## Import and tidy the amyloid data

```{r import_amyloid}
amyloid_wide_df =
  read_csv(
    "data/data_mci/mci_amyloid.csv",
    skip = 1,
    na = c("", "NA", "Na"),
    show_col_types = FALSE
  ) |>
  janitor::clean_names() |>
  rename(
    id = study_id,
    time_0 = baseline
  )

amyloid_df =
  amyloid_wide_df |>
  pivot_longer(
    cols = starts_with("time_"),
    names_to = "time",
    names_prefix = "time_",
    values_to = "amyloid_ratio",
    values_drop_na = FALSE
  ) |>
  mutate(time = as.integer(time)) |>
  arrange(id, time)

glimpse(amyloid_df)
```

I skipped the first row containing descriptions and treated both "NA" and
"Na" as missing values. I standardized column names, renamed study_id to
id, and renamed the baseline measurement column to time_0.

I reshaped the five measurement columns into a numeric time variable and
an amyloid ratio variable. Time represents years since baseline, with
measurements scheduled at 0, 2, 4, 6, and 8 years. I retained missing
measurements and sorted the data by participant ID and time.

The amyloid dataset contains `r n_distinct(amyloid_df$id)` participants.
The tidy dataset has `r nrow(amyloid_df)` rows and
`r ncol(amyloid_df)` columns, with one row per participant and scheduled
time point. There are `r sum(is.na(amyloid_df$amyloid_ratio))` missing
amyloid values and `r sum(!is.na(amyloid_df$amyloid_ratio))` observed values.

## Compare participant IDs

```{r compare_ids}
baseline_ids =
  baseline_df |>
  distinct(id)

amyloid_ids =
  amyloid_df |>
  distinct(id)

baseline_only =
  anti_join(baseline_ids, amyloid_ids, by = "id")

amyloid_only =
  anti_join(amyloid_ids, baseline_ids, by = "id")

amyloid_only_original =
  anti_join(amyloid_ids, baseline_raw, by = "id")

baseline_only
print(amyloid_only, n = Inf)
```

Comparing the cleaned datasets, `r nrow(baseline_only)` participants appear
only in the eligible baseline dataset, and `r nrow(amyloid_only)`
participants appear only in the amyloid dataset.

Of those appearing only in the amyloid dataset,
`r nrow(amyloid_only_original)` are absent from the original baseline file.
The remaining
`r nrow(amyloid_only) - nrow(amyloid_only_original)`
were present in the original baseline file but were excluded because they
already had MCI at baseline.

## Combine and export the data

```{r combine_mci}
mci_df =
  inner_join(baseline_df, amyloid_df, by = "id") |>
  arrange(id, time)

dim(mci_df)
n_distinct(mci_df$id)

write_csv(mci_df, "data/mci_combined.csv")
```

I joined the eligible baseline data to the longitudinal amyloid data by
participant ID, retaining only participants present in both datasets.

The combined dataset contains `r n_distinct(mci_df$id)` participants,
`r nrow(mci_df)` rows, and `r ncol(mci_df)` columns. Each row represents one
participant at one scheduled time point. Baseline demographic information
is repeated across that participant's five time points. The columns
include ID, baseline age, sex, education, APOE4 carrier status, age at MCI
onset, time since baseline, and amyloid ratio.

Missing amyloid measurements were retained; there are
`r sum(is.na(mci_df$amyloid_ratio))` missing amyloid values in the combined
dataset. I exported the resulting data to data/mci_combined.csv.