---
title: "P8105 Homework 2"
author: "xj2357"
output: github_document
---

```{r setup, include = FALSE}
library(tidyverse)
```

# Problem 1

## Import the subway data

```{r import_subway}
subway_df =
  read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
  show_col_types = FALSE) |>
  janitor::clean_names() |>
  select(line, station_name, station_latitude, station_longitude,
  starts_with("route"), entry, vending, entrance_type, ada) |>
  mutate(entry = entry == "YES")

glimpse(subway_df)
```
The cleaned dataset contains `r nrow(subway_df)` entrance/exit records and `r ncol(subway_df)` variables. The variables describe the subway line, station name and coordinates, routes served, entry permission, vending availability, entrance type, and ADA compliance. 
I standardized the column names, retained the requested variables, and converted entry from YES/NO to a logical variable. The route information still needs tidying because it is spread across route1 through route11 rather than stored in a single route-name column.

## Station counts and entry access

```{r subway_questions}
n_stations =
  subway_df |>
  distinct(line, station_name) |>
  nrow()

n_ada_stations =
  subway_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()

entry_prop_no_vending =
  subway_df |>
  filter(vending == "NO") |>
  pull(entry) |>
  mean()
```

Stations are identified by the combination of line and station name.
There are `r n_stations` distinct stations, of which
`r n_ada_stations` are ADA compliant.

Among entrances/exits without vending, approximately
`r round(entry_prop_no_vending * 100, 2)`% allow entry.

## Tidy the route information

```{r tidy_subway_routes}
subway_long_df =
  subway_df |>
  mutate(
    across(starts_with("route"), as.character)
  ) |>
  pivot_longer(
    cols = starts_with("route"),
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name",
    values_drop_na = TRUE
  ) |>
  mutate(
    route_number = as.integer(route_number)
  )

n_a_stations =
  subway_long_df |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()

n_a_ada_stations =
  subway_long_df |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```

I converted the route columns to character variables and reshaped them
into route_number and route_name. Missing route entries were omitted.
Each row now represents one route served by an entrance/exit record.
The resulting dataset contains `r nrow(subway_long_df)` rows and
`r ncol(subway_long_df)` columns.

There are `r n_a_stations` distinct stations serving the A train,
of which `r n_a_ada_stations` are ADA compliant.

# Problem 2

# Problem 3